%
% This is a one page description of DAP2. It's based on material extracted
% from the DAP2 RFC (NASA/ESE RFC 004). jhrg 11/03/04

\documentclass[justify]{dods-paper}

\rcsInfo $Id$

\input{../dap_boiler}

\title{The Data Access Protocol --- DAP 2.0}
\author{James Gallagher, Nathan Potter, Tom Sgouros, Steve Hankin, Glenn
  Flierl}
\date{\rcsInfoDate \\ Revision: \rcsInfoRevision}

\begin{document}

\begin{center}
\textbf{What is the Data Access Protocol?}
\end{center}

The Data Acces Protocol (DAP) is a protocol for access to data organized as
name-datatype-value tuples. It is particularly suited to accesses by a client
computer to data stored on remote (server) computers which are networked to
the client computer. The protocol has been used by the Distributed
Oceanpgraphic Data System (DODS) since 1995 and subsequently by many other
projects and groups.

While the name-datatype-value model is a nearly universal \emph{conceptual}
organization of data, the actual organization of data takes nearly as many
forms as there are individual collections because there are many different
file formats, APIs and file/directory organizations used to house data. The
DAP was designed to hide the implementation of different collections of data
behind a simple language-like interface based on the name-datatype-value
conceptual model.

The DAP and its associated software components (data servers and client
libraries) form the foundation of the National Virtual Ocean Data System
(NVODS). NVODS was developed as a system that facilitates access to
oceanographic data and data products via the Internet, freeing clients
(users) from considerations of: where the data are stored; the format or data
management structure under which they are stored; and (to a significant
degree) the size of the database. NVODS (formerly known as the 'Virtual Ocean
Data Hub' -- VODHub ) was created under a 1999 Broad Agency Announcement
issued by the National Ocean Partnership Program. The concept of the
VODHub is to be "a key element of the full community-based 'system' to
broaden and improve access to ocean data..." The resulting NVODS is also
planned for use in the Integrated Ocean Observing System.

Although the DAP was originally developed by and for the oceanographic
community it has been adopted by a number of meteorological and climate
groups as well and today is extensively used in all three communities -
climate, oceanography and meteorology. 

It is important to stress the discipline neutrality of the DAP and the
relationship between this and adoption of the DAP in disciplines other than
the Earth sciences. First, because the DAP is agnostic as relates to
discipline, it can be used across the very broad range of data types
encountered in oceanography - biological, chemical, physical and geological.
But of particular interest here is that there is nothing that constrains the
use of the DAP to the Earth sciences. For example, groups in the solar
physics community have adopted the DAP for their use and proposals are under
consideration in other areas of space physics. By standardizing the DAP for
the Earth sciences we hope that this will provide an impetus for other
disciplines to adopt it as well.

Acting on a request from the NASA/ESE Standards Process Group, the DAP
version 2 was described in RFC document and submitted to the SPG. Its current
(\rcsInfoDate) status is 'Draft Community Standard'.

\end{document}

% $Log: one_page_dap.tex,v $
% Revision 1.2  2004/11/03 22:49:52  jimg
% Added date in last paragraph.
%
% Revision 1.1  2004/11/03 22:44:58  jimg
% Created using text from the introduction of the DAP2 RFC.
%