First and foremost, we want to thank all the presenters and participants
who undertook major (sometimes heroic, not always successful) efforts
to attend the meeting.  Not one of the participants or prospective
participants was unaffected by the blizzard on 18-20 April 2003.

The meeting could not have taken place without the organizational
and administrative efforts of Sherri Corpuz and Linda Miller.
They provided continual prodding and cajoling while making the
site arrangements under trying last-minute circumstances.  It has
been a real pleasure to work with them.  We thank the audio-visual
staff for providing the excellent facilities and quick responses
to our needs as they arose.

We thank the leaders of the breakout sessions, who put together
the goals and contributed the reports, and we thank the people
who took the notes that appear here-in.  

We thank the presenters who have provided their presentations for
inclusion on the web site.

We thank the staff of the Marriot Courtyard in Boulder for providing
meeting rooms and food for the first day of the meeting on
extremely short notice, when UCAR was closed.

Finally we thank the entire UCAR/Unidata staff for providing the meeting
space at the Center Green campus, and the support necessary to 
make the meeting a success.

% $Id$
%
% $Log: acknowledgements.tex,v $
% Revision 1.2  2003/05/05 16:40:45  tom
% edited, LaTeX-ized
%
% Revision 1.1  2003/04/29 14:09:43  paul
% Starting OPeNDAP Technical Working Conference 2003 Report
%
%

%%% Local Variables: 
%%% mode: plain-tex
%%% TeX-master: "acknowledgements"
%%% End: 
