% Preface to the DODS Data Standards Guide
%
% $Id$
%
% $Log: preface.tex,v $
% Revision 1.2  1999/07/01 16:05:27  tom
% added example chapter
%
% Revision 1.1  1999/06/07 05:09:32  tom
% added to CVS
%
%

\T\chapter*{Preface}
\T\addcontentsline{toc}{chapter}{Preface}

This document describes the various data standards that may be used by
the \DODS\ (DODS) data providers to facilitate the interoperability of
DODS datasets.  These standards are not essential to the operation of
the DODS software, but are only meant as a guide to those providers
for whom interoperability is an important consideration.

This document is relevant to version \DODSversion\ and later of DODS.

\W\htmlmenu{4}
\W\chapter*{Preface}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Who is this Guide for?}
\label{pref,who}

This guide is designed to address the concerns of scientists and
others who wish to use DODS to make their data available to one, two,
or zillions of other scientists (and other others).  Making data
available using DODS is usually only half of the battle.  It is often
the case that for data to be actually useful to others, it must
conform to certain data standards expected by other users or other
users' software.  This manual documents standards used and
\emph{recommended} (not required) by the DODS group.

This guide assumes a basic familiarity with the DODS software
package.  Refer to the \DODSuser\ for an overview of the package.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Organization of this Document}

\begin{description}

\item[\bf\chapterref{intro}] provides an overview of the need for
  standards. 
  
\item[\bf\chapterref{dods-standard}] describes one set of optional (and
  evolving) DODS data standards.

\item[\bf\chapterref{adding-attr}] shows an example: Making a simple
  dataset DODS compliant.

\item[\bf\appref{std-names}] contains a list of the data variable
  names currently incorporated into the DODS meta-data standard.

\end{description}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\listconventions

\tbd{Are there any other conventions that need illustration?}


%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% TeX-master: t
%%% End: 
