Jonathan Callahan reviewed the status of the \ac{LAS}.  The goals of 
\ac{LAS} include:

\begin{itemize}
   \item Web based, common interface to diverse data
   \item Single interface for subsetting, download, visualization, comparison
   \item Easy access to metadata and documentation
   \item Unified access to distributed data holdings
   \item Uniform user interface to existing back end visualization packages
\end{itemize}

He reviewed the architecture and then how that architecture is used
to deliver \ac{OPeNDAP} data.  He then went into the details of metadata
ingestion, and how \ac{LAS} provides a uniform interface to disparate
datasets.  He stressed the requirement for a common complete subset
of necessary metadata among varied datasets to make them uniformly
accessable, not only via \ac{LAS}, but by any tool that would treat
datasets uniformly with common use metadata.  He gave three examples
of \ac{NVODS} datasets in \ac{LAS}.  He followed that with an example of
a dataset configuration XML.  The \ac{AIS} is needed by \ac{LAS} because

\begin{itemize}
   \item Fix datasets by adding 'use' and 'presentation' metadata.
   \item Source files don't need to be changed.
   \item \ac{AIS} files created by users can be shared with data providers.
\end{itemize}

He then described the need for conventions for use metadata, and he listed
some of the existing conventions that one might follow.  

Finally, his recommended list of most useful 'use' metadata would include:

\begin{enumerate}
   \item convention that is being followed 
   \item variable title 
   \item variable units that follow some convention (not 'Temp' or 'YYYYMMDD') 
   \item variable bad flag 
   \item axis units 
   \item time axis encoding should follow some convention 
\end{enumerate}

% $Id$
%
% $Log: callahansummary.tex,v $
% Revision 1.2  2003/05/05 16:40:45  tom
% edited, LaTeX-ized
%
% Revision 1.1  2003/04/29 14:09:01  paul
% Starting OPeNDAP Technical Working Conference 2003 Report
%
%

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End: 
