The OPeNDAP Data Connector (ODC) provides a graphical user interface to search 
for and retrieve datasets published by OPeNDAP data servers. OPeNDAP servers 
(also known as DODS servers) located at major institutions around the world serve a 
wide variety of data including: climatic data, satellite imagery, and ocean sensor 
results. The ODC allows you to find these datasets, download them to your machine, 
save them, and import them into client applications like IDL, Matlab, SPSS, Excel, or 
into databases such as Access and Oracle. The ODC also has simple image viewing 
and data plotting capabilities.

(see presentation screenshots)

Summary of discussion questions:

\begin{enumerate}

  \item Can the spatial-temporal panel be used separately?

Yes, the code in the ODC is modular and some portions such as the Spatial-
Temporal Panel can be taken out and used independently.

  \item Can the user reconfigure the interface to hide certain panels?

No. This feature may be added at some point in the future.

  \item Can the ODC be tightly integrated with GrADS?

No. This is because GrADS uses a closed command line interface with no 
programmatic API.

   \item How can other programs interoperate with the ODC?

The ODC has a built-in command server (IPS) which allows other programs to give it 
commands and receive responses. Currently there are clients written in Excel and 
Matlab which can talk to the IPS. The ODC code base includes an example class 
written in Java which illustrates a typical IPS client.


\end{enumerate}


% $Id$
%
% $Log: chamberlainsummary.tex,v $
% Revision 1.2  2003/05/05 16:40:45  tom
% edited, LaTeX-ized
%
% Revision 1.1  2003/04/29 14:09:01  paul
% Starting OPeNDAP Technical Working Conference 2003 Report
%
%

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End: 
