Don Denbo was unable to attend due to the blizzard.  However, his
report on the Climate Data Portal is available on the web:
\xlink{http://www.epic.noaa.gov/talks/dwd/NVODS\_Workshop\_2003/ncBrowse\_CDP.htm
}{http://www.epic.noaa.gov/talks/dwd/NVODS\_Workshop\_2003/ncBrowse\_CDP.htm
} and is summarized here.  Also, he attended the in-situ breakout via
speaker-phone. 

The Climate Data Portal at PMEL is available for accessing distributed
in-situ data.  First LDAP Directory Service is used to find relevant
servers.  Second, a coarse latitude/longitude/depth/time selection is
made.  Finally, the selection is refined to individual stations using
ndEdit.  The selected data then can be downloaded and saved or displayed.
The requirements to support multi-tiered data selection are: 

\begin{enumerate}
  \item Determine what datasets are available, 
  \item Select a dataset and, given a coarse selection, get a list of 
       stations and appropriate metadata, and 
  \item retrieve stations from a list of selected stations.  
\end{enumerate} 

ncBrowse is a general
purpose, interactive tool for graphical browsing of data and attributes in 
NetCDF files.  It has OPeNDAP data access to datasets compatible with
the Java-NetCDF API, and currently is being enhanced to support 
OPeNDAP datasets with more complex data models than NetCDF supports.
Several proposals were made to enhance ease-of-use of OPeNDAP datasets
through ncBrowse: 

\begin{enumerate}
  \item Combine the DDS and DAS, 
  \item Provide an easy-to-use interface to OPeNDAP Sequence and 
    Structure objects.  
\end{enumerate} 

Specific issues listed with respect to ncBrowse included:

\begin{enumerate}
  \item DAS and DDS structures don't match
  \item DAS and DDS namespaces are not always compatible
  \item It's not always possible to matich attributes to variables
        with an automatic system.
  \item There is no guaranteed way to determine that NetCDF conventions
        apply.
  \item Data structures for DAS and DDS are substantially different
        causing a little confusion and awkward programming.
\end{enumerate}

\paragraph{Denbo's reommendations}

Develop an OPeNDAP conventions framework:

\begin{enumerate}
   \item  Formalize existing OPeNDAP convention for NetCDF files.
   \item  Develop an in-situ data access convention.
   \item  Develop a Java-OPeNDAP in-situ API.
\end{enumerate}

% $Id$
%
% $Log: denbosummary.tex,v $
% Revision 1.2  2003/05/05 16:40:45  tom
% edited, LaTeX-ized
%
% Revision 1.1  2003/04/29 14:09:01  paul
% Starting OPeNDAP Technical Working Conference 2003 Report
%
%

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End: 
