\subsubsection{Members of Group:}

Ken Casey - NOAA NODC - NODC wants to increase the amount of data they
make available to GIS users. Now archiving Marine Sanctuary Data that
needs to be available for GIS systems. Also interested in educational
applications.
 
John Caron - Unidata - Working on adding georeferencing to NetCDF
Libraries and Open GIS compliant tools.
 
Dave Brown - NCAR - Scientific programmer interested in web-based
applications.
 
Jeremy Cothran, Dave White - Univ. of South Carolina COOPS - Interested
in hurricanes and storm surge models. Exploring Minnesota Map Server and
MapLab for serving maps on the web. Interested in Open Source
approaches.
 
Steve Hankin - NOAA PMEL - Interested in GeoTIFF output from LAS.
 
John Cartwright, Ted Habermann, Roland Schweitzer - NOAA NGDC -
Developers of OPeNDAP - GIS connection.
 
\subsubsection{Partnerships for Developed Tools:}

\begin{description}
\item[ArcIMS OpeNDAP Connector:] NODC

\item[OPeNDAP to GeoTIFF Server:] NODC, Unidata, PMEL (add capability to LAS)

\item[ArcIMS, OGC WMS back ends for LAS:] PMEL
\end{description}

Roland and John will merge GeoTIFF / Web Coverage Server efforts for
next generation of GeoTIFF generation.
 
\subsubsection{UCAR and ESRI Users Conference:}

John Caron reported on a meeting between ESRI and UCAR where getting
UCAR data into a demonstration for the ESRI Users Conference next summer
was discussed. ESRI seems interested in including science data in an
ArcGlobe demonstration. Steve Kopp, the lead raster developer from ESRI,
is working on this from the ESRI side.
 
\subsubsection{Datasets to be served as ArcIMS backgrounds / GeoTIFFs:}

\begin{itemize}
\item NGDC Coastal Bathymetry
\item NESDIS Satellite Products
\item NODC Satellite Products
\end{itemize}

Some questions about vectors / points in OPeNDAP. In-situ standards may
be helpful there.
 
 
% $Id$
%
% $Log: gisbreakout.tex,v $
% Revision 1.2  2003/05/05 16:40:45  tom
% edited, LaTeX-ized
%
% Revision 1.1  2003/04/29 14:09:01  paul
% Starting OPeNDAP Technical Working Conference 2003 Report
%
%

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End: 
