In order to make datasets easily accessible to clients in a uniform
manner, they should include the following use metadata at a minimum:

\begin{enumerate}
   \item The convention being followed,
   \item the variable title,
   \item the variable units that follow some convention,
   \item the variable "bad" flag,
   \item the axis units,
   \item the time axis encoding should follow some convention.
\end{enumerate}

Server loading needs to be studied and controlled based on either
or both of:  volume of large requests and/or number of requests
per unit time.

End-to-end projects that use OPeNDAP as the data transport mechanism
can work wonders for existing systems where the initial system originally
relied on hand-to-mouth 19th century technologies for data transport.

Large scale projects can and are successfully using the OPeNDAP
transport mechanism for moving large volumes of data within
the projects.  The use of OPeNDAP mechanism is becoming critical
to the day-to-day operation of more and more projects.

Several up-and-coming technologies,including Anagram, the \ac{AIS},
and an aggregation server for in-situ data, will help meet the
current and future needs of modelers and the \ac{ESG}.  

New instrument technologies are making the routine collection of
data automatic.  The storage and transmission of those data with
the appropriate metadata often is problematic.  The solution may
be to build in metadata generators into the instrumental transmission/
collection systems, and put the data into a ``standard'' format such
as netCDF.

Data access speed and efficiencies depend on server/connection/client
configurations, and whether or not the server and client are local
to each other.  The efficiency also depends on the mode of access.
Studies need to sort out the differences of the efficiencies including
all these and other factors, in order to better determine which
systems work best in a given situation.  A particular system may not
be efficiently used if the capabilities of that system are not fully
understood by the individual developer.

% $Id$
%
% $Log: servingandaccessingconclusion.tex,v $
% Revision 1.2  2003/05/05 16:40:45  tom
% edited, LaTeX-ized
%
% Revision 1.1  2003/04/29 14:09:01  paul
% Starting OPeNDAP Technical Working Conference 2003 Report
%
%

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End: 
