Phil Sharfstein could not attend the meeting because of the blizzard, but
his presentation was given by Steve Hankin.  

The \ac{USGODAE} Monterey Data Server has been established to support
the international \ac{GODAE} program.  It hosts in-situ data, atmospheric
forcing fields suitable for driving ocean models, and unique \ac{GODAE}
datasets including demonstration models.  The server hardware and
software are designed to be robust against interruptions to 24x7
data flow.  Supported web site and server applications include:

\begin{itemize}
  \item Consistent, easy-to-navigate web site
  \item Data browser
  \item Observation file thumbnail generator creates preview plots
        of non-gridded data.
  \item Argo \ac{GDAC} and web front-end
  \item \ac{LDM}
  \item \ac{LAS}
  \item \ac{OPeNDAP} Servers
  \item \ac{THREDDS} Catalogs
\end{itemize}

\ac{OPeNDAP} Servers in use are: \ac{GDS}, standard \ac{OPeNDAP} 
Servers, and \ac{OPeNDAP} Aggregation Servers.  The U.S. \ac{GODAE}
role has been expanded to include serving the needs of \ac{CONDUIT}.
The report gives a comprehensive review of the large volume and
variety of datasets available from the Monterey Data Server.  
Current GODAE projects include:

\begin{itemize}
  \item Hosting data and applications for \ac{GODAE} projects
  \item \ac{GODAE} Modeler's LAS
  \item Ocean QC comparison
  \item Surface Flux Comparison
\end{itemize}

Future Developments include:

\begin{itemize}
  \item Continued development of the \ac{USGODAE} site to create a one-stop
        source for \ac{USGODAE} data and information,
  \item Outreach to ocean modeling community for integration of ocean
        models through NVODS or direct hosting on \ac{USGODAE}
  \item Continued addition of new data and ocean models from providers 
        throughout the world and integration of those data into
        \ac{LAS} and \ac{OPeNDAP}.
  \item Increased promotion of the \ac{USGODAE} server and data to the
        \ac{GODAE} community.
\end{itemize}
% $Id$
%
% $Log: sharfsteinsummary.tex,v $
% Revision 1.2  2003/05/05 16:40:45  tom
% edited, LaTeX-ized
%
% Revision 1.1  2003/04/29 14:09:01  paul
% Starting OPeNDAP Technical Working Conference 2003 Report
%
%

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End: 
