%
% Documentation for the DAP. Intended to be like an RFC document.
%

\documentclass[justify]{dods-paper}
\usepackage{longtable}
\usepackage{acronym}
\usepackage{xspace}
\usepackage{gloss}
\usepackage{changebar}
\rcsInfo $Id$

\include{dap_boiler}

% Note: to get the glossary to work, run bibtex on the *.gls.aux file,
% then latex the file, then bibtex *.gls, then latex again. Also, make
% sure to set your BST and BIBINPUTS environment variables so that the
% BST and BIB files will be found.
% \makegloss

\title{THINGS WE HAVE TO DO}
\author{Nathan Potter, James Gallagher}

\begin{document}

\maketitle

\section{Questions}
\begin{itemize}
\item Should we switch URI for URL?

\item Should we have a section about clients that cannot handle certain
  types? What should they do? What about servers that gag on something? We do
  have a quasi-standard way to handle errors in attributes implemented by the
  HDF4 server\ldots
  
\item Limit types, the way types can be combined, or keep the spec completely
  general? For example, will any client be able to process an Array of
  Sequences? Sure, we could write one, but will anybody else?
  
\item Supported relational expressions for the String type are limited to
  equality, inequality and regular expression match. The less than operations
  are just too wierd in real life.
  
\item We need to find that reference for Out of Band data (from the W3C).

\item Are we going to Base64 encode Attributes.

\item [DONE] Should the XML syntax for \Grid be changed to eliminate
the \Target element? We could do this by adding a \lit{role} attribute
to the \Alias element and specifying that the allowable values for
\lit{role} are \lit{array} and \lit{map}. The \lit{role} attribute
would only be allowed in \Aliases that are member variables of a
\Grid.\\ ANSWER: YES!

\item Should the XML syntax for \Constraint be changed to make the
\NoAttributes element (which is a really a simple switch) into an XML
attribute of \Constraint?  This might look like an option Xml
attribute called \lit{Attributes} whose value could be set to
\lit{yes} or \lit{no} with it defaulting to \lit{yes} if the attribute
was missing.\\ANSWER: Probably not...

\item In the section on the Blob we may need to be more exact about
  how lengths are sent fro Arrays.

\item in subsubsection Constructor variable names, I think the []
  notation is out of place. It would be better to adopt a
  syntax-neutral notation.

\item The sentence ``In essence all variable members variables of a
constructor variable act as \Attribute structures within the
constructor variable.'' from label{sec-FQN} is really hard to
understand.

\item In subsection Processing Attributes, we should explain how the
  origin facet of a variable can form a linkage with a modifier
  attribute.

\item We will need to spec out just what a regular expression
  is. Probably in an Appendix.

\end{itemize}

\section{Important Points}
\begin{itemize}
\item Grids now may have $N$ dimensional maps.
\item We're introducing some new atomic types: Boolean, Enumeration, Int64,
  UInt64, and Time.
\end{itemize}

\section{Dap Objects TODO}
 
\begin{description}
\item Uppercase MUST, MAY NOT, etc.\\[1mm]
[DONE] MUST \\[1mm]
[DONE] MAY \\[1mm]
[DONE] MAY NOT \\[1mm]
others?
 
\item [DONE] Add origin attribute to Dataset?

\item [DONE] Add (and name) an \Attribute free \DDX. FIX: Added tag to \CE 
that does the trick.
 
\item [DONE] There are significant conflicts in general discussion of Grid; in Variables section, Responses (DDX XML Elements) section, and in examples. 
In particular:\\[2mm]
$\bullet$ [DONE] The Maps cannot yet be Aliases. (FIX: allow aliases)\\
$\bullet$ [DONE] The discussion of the way MAP is related to a dimension in the ARRAY is ambiguous and contradictory (both, can you believe it?): Is it by dimension name? OR is it by the name of the MAP element? What about multiple dimension MAP elements? (FIX: Assign by dimension name.\emph{This is a show stopper for releasing this document.}
 
\item [DONE] Fully qualified names vs partial names in projection and selection. FIX: Make all fully qualified. (Can we start with slash and skip dataset name?)

 
\item [DONE] Sequence variables poorly defined in DDX XML Elements. Consider adding:\\[2mm] $\bullet$ Alias may not point to a member of Sequences.\\
$\bullet$ Sequences cannot contain Aliases that point outside the Sequence.
FIX: Make it so. Provide some justification. Implicit relation argument.

\item [DONE] Can you have \Arrays of \Enums? FIX: No.

\item [DONE] What about allowing selection based on Attribute content? Say, all variables with origin="helena"? All variables with a attributed named "units"? All variables with an Attribute named "units" whose value is "cm"? FIX: Yes, think about it at length. Add note to that affect

\item [DONE] dap objects document needs page numbers.
dap objects document has formatting issues (it's too long on my pages when I print it out.) FIX: Ask Tom S.

\item [DONE] Why do we restrict Attributes to being homogeneous? Why not both values and child Attributes? FIX: That's just the way it is. "Because"

\item [DONE] Naming rules, encoding rules, and length specification might all go in a separate section? What about discussing partial implementations? Example netcdf lib doesn't do sequences. FIX: Added sub-section to Responses Section.

\item Add FAQ/release notes/new features

\end{description}


\end{document}
