% Preface to the Matlab GUI User Guide
%
% $Id$
%

\T\chapter*{Preface}
\T\addcontentsline{toc}{chapter}{Preface}

\newcommand{\MGUIversion}{6.0}
\newcommand{\DAPversion}{2.0}
\newcommand{\loaddodsversion}{3.4.5}
\newcommand{\loaddapversion}{3.4.4}

This document describes version \MGUIversion\ of the \OPD\ (OPeNDAP)
Matlab Graphical User Interface (GUI), a system intended to allow
researchers transparent access to earth science data---stored in any
of several different file formats---across the Internet. The \GUI\ was
originated by Deirdre Byrne and Peter Cornillon.

This document is relevant to the following software versions:

\begin{tabular}{ll} \\
\GUI\ & \MGUIversion \\
OPeNDAP data access protocol & \DAPversion \\
loaddods (Matlab command-line client) & \loaddodsversion \\
loaddap (enhanced Matlab command-line client) & \loaddapversion 
\end{tabular}

We are interested in feedback and bug reports about the \GUI\
Please send them to: \OPDsupport .

\begin{ifhtml}
  \htmlmenu{4}
  \chapter*{Preface}
\end{ifhtml}

Using the \GUI, a researcher can browse and retrieve earth
science data from many different sources, importing it directly into
Matlab, where it is available for display or for further
manipulation.  The \GUI\ provides a graphical environment within which
a researcher can browse many datasets to find appropriate data, can
select or subsample that data, and can display it as well, all by
using a simple point-and-click user interface.

An overview of the system's use is presented, and specific tasks
illustrated, for data providers as well as for users.

This document also covers the \OPD\ Matlab command-line clients, with
which a researcher can download data directly into Matlab, for further
analysis.  The \GUI\ provides a graphical front-end to such a client,
but you can use them directly as well, which can be of use in
automating data analysis, for example.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Tasks Illustrated in this Guide}
\label{pref,tasks}

For a quick start to getting, installing, and using \GUI\
software, see the list below of tasks described in this document.

\begin{itemize}

\item Quick start. (\pagexref{gui,quickstart}) (You may also
  appreciate \OPDquick for a quick overview of the \OPD.)

\item Getting and installing the \GUI\ software. 
(\pagexref{gui,install})

\item Installing the rest of the OPeNDAP software. 
(\pagexref{gui,install})

\item Updating the dataset list to include the most recent versions of
all the datasets. (\pagexref{gui,intro,update})

\item Reading your own datasets into the browser. 
(\pagexref{gui,adding})

\item Using OPeNDAP from the \matlab\ command
  line. (\pagexref{gui,intro,loaddods}) 

% \item Troubleshooting. (\pagexref{gui,trouble})

\end{itemize}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Who is this Guide for?}
\label{pref,who}

This guide is for people who wish to use Matlab to browse and retrieve
data from existing on-line OPeNDAP datasets. Scientists who wish
to share their data with colleagues may also find this a useful
system, since it is a relatively simple matter to set up a server that
can allow remote access to your data.

This documentation assumes that the readers are familiar with
computers and the internet, but are not necessarily programmers. More
than a passing familiarity with Matlab will be useful, as will an
understanding of elementary internet concepts, such as URLs and http.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Organization of this Document}

This book contains both introductory and reference material. There is
also a description of the installation procedure.

\begin{description}

\item[\chapterref{gui,intro}] provides an overview 
of the entire system, including an introduction to its use, and some
tutorial-style examples.

\item[\chapterref{gui,ref}] Provides a reference to the menus
and operations used in the \GUI.

\item[\chapterref{gui,adding}] Describes how to add your own
datasets to the \GUI.

\item[\appref{gui,install}] Contains the instructions 
for installing the \GUI, and the related OPeNDAP software you may
require.

%\item[\appref{gui,trouble}] Presents a guide to trouble 
%shooting, in case you have trouble with the \GUI.

%  \item[\bf Glossary] A small but useful collection of terms.

\end{description}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\listconventions

%\W\section{Notices}

% $Log: preface.tex,v $
% Revision 1.8  2004/12/09 03:04:14  tomfool
% updated for GUI 6
%
% Revision 1.7  2002/02/10 02:59:47  tom
% updated for GUI version 5 and dods-book2 templates
%
% Revision 1.6  1999/05/27 15:58:39  tom
% expunged remaining references to "dods.gso.uri.edu"
%
% Revision 1.5  1999/05/25 20:47:47  tom
% modifications for DODS release 3.0
%
% Revision 1.4  1999/02/04 17:42:13  tom
% modified to use dods-book.cls and Hyperlatex
%
% Revision 1.3  1999/01/20 15:18:39  tom
% updated, and changed for dods-book.cls and hyperlatex
%
% Revision 1.2  1998/12/07 15:41:15  tom
% updated for DODS v2.19 and GUIv0.7
%
% Revision 1.1  1998/02/12 18:10:42  tom
% added to CVS archive
%
%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End: 
